---
title: "13 APIs I"
logo: "../images/mac.png"
---


## JSON files

JSON (JavaScript Object Notation) is a lightweight, human-readable data-interchange format.

- It is easy for humans to read and write.
- It is easy for machines to parse and generate.
- It is commonly used for APIs (Application Program Interface) to share data.


Here is an example of the most basic format of JSON:

```json
{
  "name": "John",
  "age": 30,
  "city": "New York",
  "hobbies": ["reading", "hiking", "coding"]
}
```


JSON Syntax:

- Curly braces `{}` enclose an object, which is a collection of key-value pairs.
- Key-value pairs--each pair consists of a key (string) and a value (can be string, number, boolean, array, or another object).
- Colon `:` separates the key from the value.
- Comma `,` separates different key-value pairs within the object.
- Square brackets `[]` enclose an array, which is an ordered list of values.
- Double quotes used to surround the key and any string value


## Reading JSON Files

There are a few packages we could use to read JSON files into R.

We will start by using the `jsonlite` package today.

```{r echo=TRUE}
# install.packages('jsonlite')
library(jsonlite)

# R originally treats JSON as a string of characters
json <- '{"name": "John", "age": 30, "city": "New York", "hobbies": ["reading", "hiking", "coding"]}'
```



```{r echo=TRUE}
#fromJSON() converts JSON characters to R objects
data_in_r <- fromJSON(json)

data_in_r
```

What type of object is this?


A named list (but elements are not all of the same length).

```{r echo=TRUE}
data_in_r |> class()
```


If we have an array of objects, we can imagine each element of the array is one row of a dataset.

```{r echo=TRUE}
json <-
'[
  {"Name" : "Mario", "Age" : 32, "Occupation" : "Plumber"}, 
  {"Name" : "Peach", "Age" : 21, "Occupation" : "Princess"},
  {},
  {"Name" : "Bowser", "Occupation" : "Koopa"}
]'
```


```{r echo=TRUE}
# If we don't try to simplify, fromJSON() provides a list
fromJSON(json, simplifyVector = FALSE) 
```


```{r echo=TRUE}
# If it is an array that satisfies the constraints of a data frame, it may be able to simplify to data frame
fromJSON(json, simplifyVector = TRUE)
```

For more info about converting JSON data to R objects with jsonlite, see <https://cran.r-project.org/web/packages/jsonlite/vignettes/json-mapping.pdf>
